Supplementary Materials for cILR: Taxonomic enrichment analysis with isometric log ratios

### 2021-12-03 cbea_revised.tex
New updates

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}
The microbiome is the collection of microorganisms (bacteria, protozoa, archaea, fungi, and viruses) which co-exist with their host. Previous research has shown that changes in the composition of the human gut microbiome are associated with important health outcomes such as inflammatory bowel disease \cite{proctor2019}, type II diabetes \cite{sharma2019a}, and obesity \cite{aoun2020}. To understand the central role of the microbiome in human health, researchers have relied on high-throughput sequencing methods, either by targeting a specific representative gene (i.e. amplicon sequencing) or by profiling all the genomic content of the sample (i.e. whole-genome shotgun sequencing) \cite{cho2012}. Raw sequencing data is then processed through a variety of bioinformatic pipelines \cite{callahan2016, truong2015}, yielding various data products, one of which are taxonomic tables which can be used to study associations between members of the microbiome and an exposure or outcome of interest.

@@

The CBEA method generates sample-specific enrichment scores for microbial sets using the isometric log-ratio transformation \cite{egozcue2003}. Details on the computational implementation of CBEA can be found in the supplemental materials. The CBEA method takes two inputs:  

@@

    \item $\mathbf{X}$: $n$ by $p$ matrix of positive counts for $p$ taxa and $n$ samples measured through either targeted sequencing (such as of the 16S rRNA gene) or whole genome shotgun sequencing. Usually $\mathbf{X}$ is generated from standard taxonomic profiling pipelines such as DADA2 \cite{callahan2016} for 16S rRNA sequencing, or MetaPhlAn2 \cite{truong2015} for whole genome shotgun sequencing. 

@@

\subsection*{Properties of CBEA}
\subsubsection*{CBEA and balances of groups of parts}
The CBEA statistic is a special instance of the isometric log-ratio transformation (ILR) \cite{egozcue2003}. The standard ILR is a transformation method to address the negative correlation bias inherent in compositional data by providing an isometry between the $D$-dimensional simplex $\mathbb{S}^D$ and coordinates in the $D-1$ real space $\mathbb{R}^{D-1}$ \cite{egozcue2003,washburne2017}. This is accomplished by projecting the composition onto a chosen orthonormal basis in $\mathbb{R}$, which can be defined by a sequential binary partition (SBP) of the variables (e.g. a rooted phylogenetic tree). The ILR transformed variables are the coordinates of nodes within an SBP tree of the variables. Without loss of generalizability, in a given SBP with node $i$ splitting variables between sets $R$ and $S$, we have the ILR coordinate $x^{*}_{i}$ as: 
\begin{equation}\label{ilr_standard}
    x^{*}_i = \sqrt{\frac{rs}{r+s}} \log\left(\frac{g(X_{j|j \in R})}{g(X_{j|j \in S})}\right)
\end{equation}
where $r$ and $s$ are the cardinalities of sets $R$ and $S$ respectively, $g(z)$ is the geometric mean, and $X_{j}$ are values of the original predictors with indexes defined by membership in $R$ and $S$. The ILR confer many important benefits. First, ILR coordinates exist in real space, whereby common statistical methods can be used. Second, ILR aggregated variables preserve inter-sample distances before and after aggregation \cite{egozcue2005}. Third, ILR variables are not constrained to sum to 0 as that of the centered log-ratio transformation, resulting in a covariance matrix that is not singular \cite{egozcue2003}.

The usage of the ILR statistic is not uncommon in the microbiome literature. They are usually termed ``compositional balances", and have been leveraged in many recent approaches in variable transformation \cite{washburne2017,silverman2017,morton2017}. The CBEA formulation in Eq~\eqref{main_eq} is a special case of Eq~\eqref{ilr_standard} defined on a node that splits the taxa into two disjoint sets, one representing the set of interest, the other representing the remaining taxa. As such, the CBEA transformation inherits the properties of the ILR as a log-ratio method applicable to compositional data sets. However, unlike the ILR and its variants \cite{silverman2017, morton2017, washburne2017}, the axes defined by each CBEA set are not orthogonal (since the balances are mutually exclusive between sets and do not belong in the same SBP). Hence, a correlation can exist between CBEA aggregated variables.

\subsubsection*{Estimating the null distribution}
We can perform significance testing on the CBEA statistic which corresponds to the null hypothesis that the center of the subcomposition defined by the set is equal to the center of the subcomposition defined by the complement of the set. This is equivalent to the $Q_1$ competitive null hypothesis in the gene set testing literature \cite{tian2005} where the enrichment of a gene set is defined with respect to genes outside the set.

We can apply prior usage of the ILR statistic in hypothesis testing to CBEA by assuming that the null distribution of CBEA follows a standard normal distribution \cite{egozcue2005}. However, when applying CBEA for hypothesis testing at the sample level, it is expected that the researcher would be testing a large number of hypotheses. Under the assumption that the number of truly significant hypotheses is low, Efron \cite{efron2004} showed that estimating the null distribution of the test statistic (termed the empirical null distribution) is much more preferable than using the theoretical null due to unobserved confounding effects inherently part of observational studies. As such, to perform significance testing using CBEA, we also estimated the null distribution from observed raw CBEA variables.

@@

We performed our empirical null estimation by fitting our distribution of choice and computing relevant parameters on raw CBEA scores on taxa-permuted data (equivalent to gene permutation in the gene expression literature). As such, the null distribution is characterized by scores computed on sets of equal size with randomly drawn taxa. However, null distribution based on taxa-permutation is sensitive to inter-taxa correlations within the set \cite{wu2012}. Since the permutation procedure does not preserve correlation structures, estimating parameters from empirical scores on permuted data will underestimate the variance inflation due to correlation. We account for this by combining the mean estimate from permuted data with the variance estimate from unpermuted data, where the inter-taxa correlation structure remains undisturbed. However, this procedure assumes that the variance of CBEA is equal under both the null and alternate hypotheses.

\subsection*{Evaluation}
All code and data sets used for evaluation of this method is publicly available and can be found on GitHub (\href{www.github.com/qpmnguyen/CBEA\_analysis}{qpmnguyen/CBEA\_analysis}).

\subsubsection*{Parametric Simulations}  
To address the performance of CBEA for different modeling tasks, we simulated microbiome count data under the assumption that it follows a zero-inflated negative binomial distribution, which is a good fit for real microbiome relative abundance data \cite{calgaro2020}. Suppose $X_{ij}$ are observed counts for a sample $i$ and taxon $j$, then we have the following probability model
\begin{equation}
    \mathbf{X}_{ij} =
      \begin{cases}
        0 & \text{with probability $p_j$}\\
        \mathbf{NB}(\mu_j, \phi_j) & \text{with probability $1 - p_j$}\\
      \end{cases}       
\end{equation}

@@

\noindent \textbf{Single Sample Enrichment}: To assess type I error rate and power for enrichment significance testing at the sample level, we simulated data based on the schema above, and assessed enrichment for one focal set. Type I error was obtained under the global null as the number of samples where the null hypothesis was rejected at $\alpha = 0.05$ over the total number of samples (which represents the total number of hypotheses tested). Power was obtained using the same formulation as type I error rate but under the global alternate. We treated type I error and power as estimates of binomial proportions and utilized the Agresti-Couli \cite{agresti1998} formulation to calculate 95\% confidence intervals. Across both analyses, we varied sparsity levels ($p = 0.2, 0.4, 0.6$) and inter-taxa correlation within the set ($\rho = 0, 0.2, 0.5$). For type I error analysis, we also varied the size of the set (50, 100, 150). For power analyses, set size was kept constant at 100 but different effect sizes (fold change of 1.5, 2, and 3). All sample sizes were set at 10,000.

For classifiability, we evaluated the scores against the true labels per sample (indicating the sample has a set with inflated counts) using the area under the receiving operator curve (AUROC/AUC). This is a strategy used in Frost \cite{frost2020a} which evaluates the informativeness of scores by assessing the relative ranking of samples (i.e. whether samples with inflated counts are highly ranked using estimated scores).  DeLong 95\% confidence intervals for AUC \cite{delong1988} were obtained for each estimate. Simulation settings for classification performance were identical to power analyses as detailed in the previous paragraph.

\noindent \textbf{Differential Abundance Analysis}: To assess type I error rate and power for differential abundance testing task, we simulated data based on the schema above, and assessed differential abundance of 50 sets with 100 taxa per set across 20 replicates per simulation condition. Type I error is calculated as the number of differentially abundant sets over the total number of sets for each simulation under the global null. Power is defined similarly, but instead under the global alternate hypothesis. Estimates and confidence intervals for type I error and power are calculated as cross-replicate mean and standard error. A set is differentially abundant when all taxa within a set are differentially abundant with the same effect size. Across both analyses, we varied sparsity levels ($p = 0.2, 0.4, 0.6$), and inter-taxa correlation within the set ($\rho = 0, 0.2, 0.5$). Half of the sets are differentially abundant across case/control status with varying effect sizes (fold change of 1.5, 2, and 3). Due to the compositional nature of microbiome taxonomic data, simple inflation of raw counts would cause an artificial decrease in the abundance of the remaining un-inflated sets. As such, we applied a compensation procedure as described in Hawinkel et al. \cite{hawinkel2019} to ensure the validity of simulation results. All sample sizes were set at 2,000.

\noindent \textbf{Prediction}: To assess predictive performance, we generated predictors based on the simulation schema presented above and evaluated prediction for both binary and continuous outcomes using a standard random forest model \cite{breiman2001}. For binary outcomes, we use AUC similar to the classification analyses above. For continuous outcomes, we used root mean squared error (RMSE). All predictive model fitting was performed using \emph{tidymodels} \cite{kuhn2020} suite of packages. Across both learning tasks, we varied sparsity ($p = 0.2, 0.4, 0.6$), and inter-taxa correlation ($\rho = 0, 0.2, 0.5$). Continuous outcomes $Y_{cont}$ were generated as linear combinations of taxa counts.  
\begin{equation}
    Y_{cont} = f(\mathbf{X}) + \mathbf{\epsilon}
\end{equation}
where $\mathbf{\epsilon} \sim N(0, \sigma_{\epsilon}^2)$ and $f(\mathbf{X}) = \beta_0 + \mathbf{X}\mathbf{\beta}$. For each simulation, we set $\beta_0$ to be $\frac{6}{\sqrt{10}}$ similar to \cite{xiao2018}. The degree of model saturation (the number of non zero $\mathbf{\beta}$ values) were varied between 0.1 and 0.5, and signal to noise ratio (SNR = $\frac{\sigma(f(\mathbf{X}))}{\sigma_{\epsilon}}$) was varied between 1.5, 2, and 3.

@@

\subsubsection*{Real Datasets}

In addition to simulation analyses, we also evaluated our method using real data sets based on both 16S rRNA gene sequencing and whole-genome sequencing. All data sets are obtained from either the \emph{curatedMetagenomicData} \cite{pasolli2017} and \emph{HMP16SData} \cite{schiffer2019} R packages (2020-10-02 snapshot), or downloaded from the Qiita platform \cite{gonzalez2018}.

@@

\bibliography{mac_taxagg}{}

>>>>

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}
The microbiome is the collection of microorganisms (bacteria, protozoa, archaea, fungi, and viruses) which co-exist with their host. Previous research has shown that changes in the composition of the human gut microbiome are associated with important health outcomes such as inflammatory bowel disease \cite{proctor2019}, type II diabetes \cite{sharma2019}, and obesity \cite{aoun2020}. To understand the central role of the microbiome in human health, researchers have relied on high-throughput sequencing methods, either by targeting a specific representative gene (i.e. amplicon sequencing) or by profiling all the genomic content of the sample (i.e. whole-genome shotgun sequencing) \cite{cho2012}. Raw sequencing data is then processed through a variety of bioinformatic pipelines \cite{callahan2016, truong2015}, yielding various data products, one of which are taxonomic tables which can be used to study associations between members of the microbiome and an exposure or outcome of interest.

@@

The CBEA method generates sample-specific enrichment scores for microbial sets using products of proportions \cite{egozcue2003}. Details on the computational implementation of CBEA can be found in the supplemental materials. The CBEA method takes two inputs:  

@@

    \item $\mathbf{X}$: $n$ by $p$ matrix of positive counts for $p$ taxa and $n$ samples measured through either targeted sequencing (such as of the 16S rRNA gene) or whole genome shotgun sequencing. Usually $\mathbf{X}$ is generated from standard taxonomic profiling pipelines such as DADA2 \cite{callahan2016} for 16S rRNA sequencing, or MetaPhlAn2 \cite{truong2015} for whole genome shotgun sequencing. CBEA does not accept $X$ matrices with zeroes since it invalidates the log-ratio transformation. Users can generate a dense matrix $X$ using a method of choice, however the default mode for CBEA would be add a pseudocount of 1 if zeroes are detected in the matrix. 

@@

\subsection*{Properties of CBEA}
\subsubsection*{CBEA and balances of groups of parts}
The CBEA statistic is based on the multiplication-based aggregation approach used to calculate balances between groups of parts \cite{egozcue2005}. These balances are computed using the isometric log ratio (ILR) transformation \cite{egozcue2003} formula. For a given balance $i$ splitting variables across sets $R$ and and $S$, we have the balance coordinate $x^{*}_{i}$ as: 
\begin{equation}\label{ilr_standard}
    x^{*}_i = \sqrt{\frac{rs}{r+s}} \log\left(\frac{g(X_{j|j \in R})}{g(X_{j|j \in S})}\right)
\end{equation}
where $r$ and $s$ are the cardinalities of sets $R$ and $S$ respectively, $g(z)$ is the geometric mean, and $X_{j}$ are values of the original predictors with indexes defined by membership in $R$ and $S$.

CBEA belongs to a set of methods that seeks to leverage compositional balances for the analysis of microbiome data \cite{washburne2017, rivera-pinto, silverman2017, morton2017}. Unlike methods such as PhILR \cite{silverman2017}, CBEA does not provide a sequential binary partition that forms the basis for ILR procedure \cite{egozcue2003} and is therefore not a subclass of ILR. A similar method to CBEA would be phylofactor \cite{washburne2017}. However, instead of performing an optimization procedure to identify interesting balances, CBEA constructs balances apriori using pre-defined sets, and formulates the enrichment of a set as the scaled log-ratio between the center of the subcomposition represented by microbes within the set and the center of the subcomposition represented by remainder taxa. This formulation aligns with the $Q_1$ null hypothesis from the gene set testing literature \cite{tian2005}.

\subsubsection*{Estimating the null distribution}
We can assume that the CBEA statistic, similar to other log-ratio based transforms, follows a normal distribution \cite{egozcue2003, aitchison1980}. However, when applying CBEA for hypothesis testing at the sample level, it is expected that the researcher would be testing a large number of hypotheses. Under the assumption that the number of truly significant hypotheses is low, Efron \cite{efron2004} showed that estimating the null distribution of the test statistic (termed the empirical null distribution) is much more preferable than using the theoretical null due to unobserved confounding effects inherently part of observational studies. As such, to perform significance testing using CBEA, we also estimated the null distribution from observed raw CBEA variables.

@@

We performed our empirical null estimation by fitting our distribution of choice and computing relevant parameters on raw CBEA scores on taxa-permuted data (equivalent to gene permutation in the gene expression literature). As such, the null distribution is characterized by scores computed on sets of equal size with randomly drawn taxa.

However, null distribution based on taxa-permutation is sensitive to inter-taxa correlations within the set \cite{wu2012}. Since the permutation procedure does not preserve correlation structures, estimating parameters from empirical scores on permuted data will underestimate the variance inflation due to correlation. We account for this by combining the mean estimate from permuted data with the variance estimate from unpermuted data, where the inter-taxa correlation structure remains undisturbed. However, this procedure assumes that the variance of CBEA is equal under both the null and alternate hypotheses.

\section*{Evaluation}
All code and data sets used for evaluation of this method is publicly available and can be found on GitHub (\href{www.github.com/qpmnguyen/CBEA\_analysis}{qpmnguyen/CBEA\_analysis}).

\subsection*{Simulation Analysis}
\subsubsection*{Data generation model}  
To address the performance of CBEA for different modeling tasks, we simulated microbiome count data under the assumption that it follows a zero-inflated negative binomial distribution, which is a good fit for real microbiome relative abundance data \cite{calgaro2020}. Suppose $X_{ij}$ are observed counts for a sample $i$ and taxon $j$, then we have the following probability model
\begin{equation}
    \mathbf{X}_{ij} =
      \begin{cases}
        0 & \text{with probability $p_j$}\\
        \mathbf{NB}(\mu_j, \phi_j) & \text{with probability $1 - p_j$}\\
      \end{cases}       
\end{equation}

@@

\subsubsection*{Simulation scenarios for enrichment analysis at the sample level}
To assess type I error rate and power for enrichment significance testing at the sample level, we simulated data based on the schema above, and assessed enrichment for one focal set. Type I error was obtained under the global null as the number of samples where the null hypothesis was rejected at $\alpha = 0.05$ over the total number of samples (which represents the total number of hypotheses tested). Power was obtained using the same formulation as type I error rate but under the global alternate. We treated type I error and power as estimates of binomial proportions and utilized the Agresti-Couli \cite{agresti1998} formulation to calculate 95\% confidence intervals. Across both analyses, we varied sparsity levels ($p = 0.2, 0.4, 0.6$) and inter-taxa correlation within the set ($\rho = 0, 0.2, 0.5$). For type I error analysis, we also varied the size of the set (50, 100, 150). For power analyses, set size was kept constant at 100 but different effect sizes (fold change of 1.5, 2, and 3). All sample sizes were set at 10,000.

For classifiability, we evaluated the scores against the true labels per sample (indicating the sample has a set with inflated counts) using the area under the receiving operator curve (AUC). This is a strategy used in Frost \cite{frost2020a} which evaluates the informativeness of scores by assessing the relative ranking of samples (i.e. whether samples with inflated counts are highly ranked using estimated scores).  DeLong 95\% confidence intervals for AUC \cite{delong1988} were obtained for each estimate. Simulation settings for classification performance were identical to power analyses as detailed in the previous paragraph.

\subsubsection*{Simulation scenarios for enrichment analysis at the population level}
To assess type I error rate and power for differential abundance testing task, we simulated data based on the schema above, and assessed differential abundance of 50 sets with 100 taxa per set across 20 replicates per simulation condition. Type I error is calculated as the number of differentially abundant sets over the total number of sets for each simulation under the global null. Power is defined similarly, but instead under the global alternate hypothesis. Estimates and confidence intervals for type I error and power are calculated as cross-replicate mean and standard error. A set is differentially abundant when all taxa within a set are differentially abundant with the same effect size. Across both analyses, we varied sparsity levels ($p = 0.2, 0.4, 0.6$), and inter-taxa correlation within the set ($\rho = 0, 0.2, 0.5$). Half of the sets are differentially abundant across case/control status with varying effect sizes (fold change of 1.5, 2, and 3). Due to the compositional nature of microbiome taxonomic data, simple inflation of raw counts would cause an artificial decrease in the abundance of the remaining un-inflated sets. As such, we applied a compensation procedure as described in Hawinkel et al. \cite{hawinkel2019} to ensure the validity of simulation results. All sample sizes were set at 2,000.

\subsubsection*{Simulation scenarios for downstream prediction}
To assess predictive performance, we generated predictors based on the simulation schema presented above and evaluated prediction for both binary and continuous outcomes using a standard random forest model \cite{breiman2001}. For binary outcomes, we use AUC similar to the classification analyses above. For continuous outcomes, we used root mean squared error (RMSE). All predictive model fitting was performed using \emph{tidymodels} \cite{kuhn2020} suite of packages. Across both learning tasks, we varied sparsity ($p = 0.2, 0.4, 0.6$), and inter-taxa correlation ($\rho = 0, 0.2, 0.5$). Continuous outcomes $Y_{cont}$ were generated as linear combinations of taxa counts.  
\begin{equation}
    Y_{cont} = f(\mathbf{X}) + \mathbf{\epsilon}
\end{equation}
where $\mathbf{\epsilon} \sim N(0, \sigma_{\epsilon}^2)$ and $f(\mathbf{X}) = \beta_0 + \mathbf{X}\mathbf{\beta}$. For each simulation, we set $\beta_0$ to be $\frac{6}{\sqrt{10}}$ similar to \cite{xiao2018}. The degree of model saturation (the number of non zero $\mathbf{\beta}$ values) were varied between 0.1 and 0.5, and signal to noise ratio (SNR = $\frac{\sigma(f(\mathbf{X}))}{\sigma_{\epsilon}}$) was varied between 1.5, 2, and 3.

@@

\subsection*{Evaluation using real data}
In addition to simulation analyses, we also evaluated our method using real data sets based on both 16S rRNA gene sequencing and whole-genome sequencing. We follow the benchmark set by Geistlinger et al. \cite{geistlinger} All data sets are obtained from either the \emph{curatedMetagenomicData} \cite{pasolli2017} and \emph{HMP16SData} \cite{schiffer2019} R packages (2020-10-02 snapshot), or downloaded from the Qiita platform \cite{gonzalez2018}.

@@

\bibliography{references}{}
